Evaluating Cloud-Optimized HDF5 for NASA’s ICESat-2 Mission

### 2024-10-03 paper.qmd
another incremental update, maybe we can finish this next week

As a result of community feedback and “hack weeks” organized by NSIDC and UW eScience Institute in 2023[@h5cloud2023], NSIDC started the Cloud Optimized Format Investigation (COFI) project to improve access to HDF5 from the ICESat-2 mission. A spaceborne lidar that retrieves surface topography of the Earth’s ice sheets, land and oceans [@NEUMANN2019111325].  Because of its complexity, large size and importance for cryospheric studies we targeted the ATL03 dataset. ATL03 core data are geolocated photon heights from the ICESat-2 ATLAS instrument. Each file contains 1003 geophysical variables in 6 data groups. Although our research was focused on this dataset, most of our findings are applicable to any dataset stored in HDF5 and NetCDF4.

@@

We tested access times to original and different configurations of cloud-optimized HDF5 ATL03 files [list files tested] stored in AWS S3 buckets in region us-west-2, the region hosting NASA’s Earthdata Cloud archives.  Files were accessed using Python tools commonly used by Earth scientists: h5py and Xarray[@Hoyer2017-su].  h5py is a Python wrapper around the HDF5 C API.  xarray^[`h5py` is a dependency of Xarray]  is a widely used Python package for working with n-dimensional data.  We also tested access times using h5coro, a python package optimized for reading HDF5 files from S3 buckets and kerchunk, a tool that creates an efficient lookup table for file chunks to allow performant partial reads of files.

The test files were originally cloud optimized by  “repacking” them, using a relatively new feature in the HDF5 C API called “paged aggregation”. Page aggregation does 2 things: it collects file-level metadata from datasets and stores it on dedicated metadata blocks in the file; and it forces the library to write both data and metadata using these fixed-size pages. Aggregation allows client libraries to read  file metadata with only a few requests and uses the page size used in the aggregation as the minimal request size, overriding the 1 request per chunk behavior.

@@

| prefix                 	| description                                        	| % file size increase 	| ~km per chunk 	| shape     	| page size 	| avg_chunk_size 	|
|------------------------	|----------------------------------------------------	|----------------------	|---------------	|-----------	|-----------	|----------------	|
| original               	| original file from ATL03 v6 (1gb and 7gb)          	| 0                    	| 1.5km         	| (10000,)  	| N/A       	| 35kb           	|
| original-kerchunk      	| kerchunk sidecar of the original file              	| N/A                  	| 1.5km         	| (10000,)  	| N/A       	| 35kb           	|
| page-only-4mb          	| paged-aggregated file with 4mb per page            	| ~1%                  	| 1.5km         	| (10000,)  	| 4MB       	| 35kb           	|
| page-only-8mb          	| paged-aggregated file with 4mb per pag8            	| ~1%                  	| 1.5km         	| (10000,)  	| 8MB       	| 35kb           	|
| rechunked-4mb          	| page-aggregated and bigger chunk sizes             	| ~1%                  	| 10km          	| (100000,) 	| 4MB       	| 400kb          	|
| rechunked-8mb          	| page-aggregated and bigger chunk sizes             	| ~1%                  	| 10km          	| (100000,) 	| 8MB       	| 400kb          	|
| rechunked-8mb-kerchunk 	| kerchunk sidecar of the last paged-aggregated file 	| N/A                  	| 10km          	| (10000,)  	| 8MB       	| 400kb          	|

@@

@@

Benchmarks show that cloud optimizing ATL03 files improved access times at least an order of magnitude when used with aligned I/O patterns, this is telling the library about the cloud optimization and page size.

@@

Placeholder

>>>>

As a result of community feedback and “hack weeks” organized by NSIDC and UW eScience Institute in 2023[@h5cloud2023], NSIDC started the Cloud Optimized Format Investigation (COFI) project to improve access to HDF5 from the ICESat-2 mission, a spaceborne lidar that retrieves surface topography of the Earth’s ice sheets, land and oceans [@NEUMANN2019111325].  Because of its complexity, large size and importance for cryospheric studies we targeted the ATL03 data product. The most relevant variable in ATL03 are geolocated photon heights from the ICESat-2 ATLAS instrument. Each ATL03 file contains 1003 geophysical variables in 6 data groups. Although our research was focused on this dataset, most of our findings are applicable to any dataset stored in HDF5 and NetCDF4.

@@

We tested access times to original and different configurations of cloud-optimized HDF5 [ATL03 files](https://its-live-data.s3.amazonaws.com/index.html#test-space/cloud-experiments/h5cloud/) stored in AWS S3 buckets in region us-west-2, the region hosting NASA’s Earthdata Cloud archives.  Files were accessed using Python tools commonly used by Earth scientists: h5py and Xarray[@Hoyer2017-su].  h5py is a Python wrapper around the HDF5 C API.  xarray^[`h5py` is a dependency of Xarray]  is a widely used Python package for working with n-dimensional data.  We also tested access times using h5coro, a python package optimized for reading HDF5 files from S3 buckets and kerchunk, a tool that creates an efficient lookup table for file chunks to allow performant partial reads of files.

The test files were originally cloud optimized by  “repacking” them, using a relatively new feature in the HDF5 C API called “paged aggregation”. Page aggregation does 2 things: fisrt, it collects file-level metadata from datasets and stores it on dedicated metadata blocks at the front of the file; second, it forces the library to write both data and metadata using these fixed-size pages. Aggregation allows client libraries to read  file metadata with only a few requests using the page size as a fixed request size, overriding the 1 request per chunk behavior.

@@

| prefix                 	| description                                        	| % file size increase 	| ~km per chunk 	| shape     	| page size 	| avg_chunk_size 	|
|------------------------	|----------------------------------------------------	|----------------------	|---------------	|-----------	|-----------	|----------------	|
| original               	| original file from ATL03 v6 (1gb and 7gb)          	| 0                    	| 1.5km         	| (10000,)  	| N/A       	| 35kb           	|
| original-kerchunk      	| kerchunk sidecar of the original file              	| N/A                  	| 1.5km         	| (10000,)  	| N/A       	| 35kb           	|
| page-only-4mb          	| paged-aggregated file with 4mb per page            	| ~1%                  	| 1.5km         	| (10000,)  	| 4MB       	| 35kb           	|
| page-only-8mb          	| paged-aggregated file with 4mb per pag8            	| ~1%                  	| 1.5km         	| (10000,)  	| 8MB       	| 35kb           	|
| rechunked-4mb          	| page-aggregated and bigger chunk sizes             	| ~1%                  	| 10km          	| (100000,) 	| 4MB       	| 400kb          	|
| rechunked-8mb          	| page-aggregated and bigger chunk sizes             	| ~1%                  	| 10km          	| (100000,) 	| 8MB       	| 400kb          	|
| rechunked-8mb-kerchunk 	| kerchunk sidecar of the last paged-aggregated file 	| N/A                  	| 10km          	| (100000,)  	| 8MB       	| 400kb          	|

@@

::: {#fig-4 fig-env="figure*"}

![](figures/figure-4.png)

shows that using paged aggregation alone is not a complete solution. This behavior us caused by over-reads of data now distributed in pages and the internals of HDF5 not knowing how to optimize 
the requests. This means that if we cloud optimize alone and use the same code, in some cases we'll make access to these files even slower. A very important thing to notice here is that rechunking the file, in this case using 10X bigger chunks results in a predictable 10X improvement in access times without any cloud optimization involved.
Having less chunks generates less metadata and bigger requests, in general is it recommended that chunk sizes should range between 1MB and 10MB[Add citation, S3 and HDF5] and if we have anough memory and bandwith even
bigger (Pangeo recommends up to 100MB chunks)[Add citation.]

@@

shows that performance once the I/O configuration is aligned with the chunking in the file, access times perform on par with cloud optimized access patterns like Kerchunk/Zarr. 
These numbers are from in-region execution. Out of region is considerable slower for the non cloud optimized case.

@@

In progress
